Apparent preferences for cognitive effort fade when multiple forms of effort

### 2021-05-31 Manuscript.Rmd
all edits implemented!

  **Standard models in neuroeconomics and behavioral economics formalize the effects of effort or delay on valuation in terms of discounting functions** [@Ainslie1975; @Frederick2002; @Green1994; @Kable2007; @kleinflugge2015; @Kool2010; @Westbrook2013]. Discounting functions appear to have a similar form for delay and effort [@Massar2015; @Prevost2010; @Seaman2018], although there is ongoing debate about how the shape of the discounting function might vary across types of demands [@Arulpragasam2018; @Biaaszek2017; @Chong2017; @Hartmann2013; @Klein-Flugge2016; @Kool2018; @Prevost2010] and the degree to which discount rates correlate across cost domains [@Seaman2018; @Westbrook2013]. A limitation of the discounting approach is that it does not directly probe how aspects of the decision context, such as opportunity costs, might influence cost evaluation, and constrains the understanding of behavioral demands to one in which such demands necessarily reduce the value of prospects.

  Single choice foraging paradigms, in which decision makers choose to accept or reject individually presented prospects, provide a natural way to investigate contextual influences on cost evaluation [@Constantino2015; @Garrett2020; @Krebs1977; @Mobbs2018; @Stephens1986]. Such paradigms require weighing a prospective reward against the perceived opportunity cost of time. Specifically, foragers must decide whether to spend their limited time by harvesting a currently available resource or by seeking out more profitable (or less costly) alternatives instead. This could involve, for example, deciding when to leave a patch of depleting richness **(e.g. when to quit a job that provides diminishing opportunities for growth)**, or choosing whether to pursue or forgo offers presented serially **(e.g. whether to accept your first job offer or wait for new potential ones)**. Here we focus on this latter type of foraging, traditionally called *prey selection* [@Krebs1977], as it provides a means to examine how choices to accept or reject individual prospects are influenced by contextual knowledge about other potentially obtainable alternatives in the same environment.

@@

```{r General prop & earnings plots (Fig. 2), eval = !figsEnd, fig.width = 10, fig.height = 4, fig.align = "center", fig.cap = "Overall choice behavior in Experiment 1. Left: Proportion accepted per cost. Middle: Total number of dollars earned by each group by the end of the experiment. Right: The relationship between proportion accepted and total earned. Consistent with the foraging design, participants who over and under accepted earned the least. These results suggest that people were willing to pursue cognitively effortful actions at the expense of earnings."}
# plots generated in 16.2.1. and 16.1.3.
#(propPlot <- prop1 | earnings1 | earnFitplot_btw)
#ggsave("./Images/FIG2_props.png", propPlot, device = "png", height = 4, width = 10)

@@

  Stability was defined as the maintenance of similar acceptance rates throughout the experimental session. The scatterplot in Fig. 4 (bottom-left and bottom-right) shows the overall proportion accepted by each participant before and after the midpoint of the session. Paired permutations showed no change in acceptance rates for the wait and easy groups (*p* > 0.05), but we found significantly lower acceptance rates in the second half of the session for the Physical group (*p* `r report_p(prepost_btw$Perms[1])`, Cohen's D = `r round(prepost_btw$CohenD[1], digits = 2)`) and the Cognitive group (*p* `r report_p(prepost_btw$Perms[2])`, Cohen's D = `r round(prepost_btw$CohenD[2], digits = 2)`). Fig. 5 (left) shows that the reduction in acceptance rates in effort groups over time was steeper for longer handling times, a hallmark of effortful demands [@Treadway2009]. We then replicated the mixed-effects logistic regression from the previous section separately for each half of the session, which confirmed that differences among demands were stable (Fig. 5, right). Together, the consistency and stability of behavior suggested that decision makers had a clear representation of the environment.

@@

  The original between-subjects task from Experiment 1 was modified so that participants foraged for cost types in addition to rewards. Every participant experienced all forms of cost (delay, physical effort, and cognitive effort). There were two block types, in which delay trials were interspersed with either physical-effort or cognitive-effort trials. This setup prevented participants from having to rapidly switch between response modalities across trials (i.e. keyboard and handgrip). Each block combination (physical/wait and cognitive/wait) was experienced three times by each participant in 7-minute-long, interleaved blocks, for a total session duration of 42 minutes. Half of the participants experienced a block sequence that started with the physical/wait block and the other half experienced a sequence that started with the cognitive/wait block. Timing parameters were held fixed throughout the experiment and were matched to the middle condition from Experiment 1 (10 s handling time and 6 s travel time). Participants were informed that this timing combination would be constant prior to beginning the experiment. The upcoming effort type was disclosed at the beginning of each block, and reward/cost offers were displayed at the start of each trial (e.g. “8 cents for physical effort”). Unlike the previous experiment, participants could express their decision to quit by pressing the spacebar during the offer window (not only during the handling time), in which case the travel time would begin immediately at the end of the 2 s offer window. Participants were trained in each type of demand until they reached the same criteria as in Experiment 1.

@@

  In Experiment 2, we tested decision makers' preferences for cognitive effort or physical effort relative to unfilled delay delay. We modified the prey-selection foraging task, so each participant experienced every type of demand in blocks that interleaved wait trials with either physical or cognitive effort (with a fixed 10 s handling time and 6 s travel time, to allow for comparisons between experiments). We separately analyzed wait trials from the two block types, labeling the two sets of wait trials according to the type of effort with which they were paired (“Wait-C” for cognitive, and “Wait-P” for physical). We hypothesized that 1) reward offer amounts would influence decisions in the same way as in Experiment 1; 2) interleaving effort and delay would produce higher acceptance rates for wait trials than for cognitive or physical effort trials, in contrast to the pattern observed in Experiment 1. and 3) that choice patterns would remain stable over time and consistent within trials.

@@

```{r BIC and parameter comparison (Fig. 9), eval = !figsEnd, fig.align = "center", fig.height = 5, fig.width = 7, fig.cap = "Computational modeling. A: BIC values for four plausible models of choice behavior for both experiments. Lower values indicate better fit. B: Differences in cost biases produced by each demand. Negative values suggest the Cognitive group acted as if it perceived its demand as valuable. C: Parameter comparison portrays consistent foraging behavior across experiments. D: Behavior recovery of the within-subjects adaptive behavior based on the winning model. The model predictions match the initial preference for cognitive effort, and the eventual convergence of acceptances by the end of the session."}
# modelPlot <- (bicPlot + costPlot) / allFitsPlot
# ggsave("./Images/FIG9_modelresult_part.png", modelPlot, device = "png", height = 8, width = 11)
include_graphics("./Images/FIG9_modelresults.png")

@@

  The observed choice dynamics ruled out a number of candidate explanations. For instance, if the subjective costs and benefits of effort or delay were intrinsic, we would not have expected to observe the gradual convergence in acceptances when all three forms of demand were intermixed in the same environment. This line of reasoning also speaks against a direct role of boredom in our results. The aversiveness of boredom can lead people to attempt any available activity in order to avoid it [@westgate2020; @Wilson2014]. In our paradigm, such a tendency would have been expected to lead to a persistent preference for either effortful condition. Lastly, the gradual convergence of acceptance rates across cost conditions  rules out the possibility that our sample was mostly composed of demand seekers. Some individuals express a strong need for cognition, which can bias studies of mental labor [@Cacioppo1996; @Sayali2019]. Even though the prevalence of this trait in our college-level population could have driven the preferences for cognitive effort observed in both experiments, we would have expected such individuals to maintain this preference in the face of alternatives. Another unlikely candidate is fatigue. Even though participants decreased their acceptance rates for effort in both experiments, we would have expected participants in Experiment 2 to increase engagement with wait trials, substituting wait trials for effort trials in order to preserve their earnings. Our observations are unlikely to be due to the risk of failing to successfully complete effortful trials, as participants remained accurate throughout both experiments. Finally, we cannot conclude that cognitive effort gained value from its rewarding outcomes, as has been proposed in the past [@Eisenberger1996]. This phenomenon would not have been expected to benegit a single type of demand differentially because cognitive effort, physical effort, and unfilled delay were rewarded at the same rate.

@@

```{r New General prop & earnings plots (Fig. 2), eval = figsEnd, fig.width = 10, fig.height = 4, fig.align = "center", fig.cap = "Overall choice behavior in Experiment 1. Left: Proportion accepted per cost. Middle: Total number of dollars earned by each group by the end of the experiment. Right: The relationship between proportion accepted and total earned. Consistent with the foraging design, participants who over and under accepted earned the least. These results suggest that people were willing to pursue cognitively effortful actions at the expense of earnings."}
# plots generated in 16.2.1. and 16.1.3.
#(propPlot <- prop1 | earnings1 | earnFitplot_btw)
#ggsave("./Images/FIG2_props.png", propPlot, device = "png", height = 4, width = 10)
 prop1 | earnings1 | earnFitplot_btw
```

@@

```{r New BIC and parameter comparison (Fig. 9), eval = figsEnd, fig.align = "center", fig.height = 5, fig.width = 7, fig.cap = "Computational modeling. A: BIC values for four plausible models of choice behavior for both experiments. Lower values indicate better fit. B: Differences in cost biases produced by each demand. Negative values suggest the Cognitive group acted as if it perceived its demand as valuable. C: Parameter comparison portrays consistent foraging behavior across experiments. D: Behavior recovery of the within-subjects adaptive behavior based on the winning model. The model predictions match the initial preference for cognitive effort, and the eventual convergence of acceptances by the end of the session."}
# modelPlot <- (bicPlot + costPlot) / allFitsPlot
# ggsave("./Images/FIG9_modelresult_part.png", modelPlot, device = "png", height = 8, width = 11)
include_graphics("./Images/FIG9_modelresults.png")
```

>>>>

  Standard models in neuroeconomics and behavioral economics formalize the effects of effort or delay on valuation in terms of discounting functions [@Ainslie1975; @Frederick2002; @Green1994; @Kable2007; @kleinflugge2015; @Kool2010; @Westbrook2013]. Discounting functions appear to have a similar form for delay and effort [@Massar2015; @Prevost2010; @Seaman2018], although there is ongoing debate about how the shape of the discounting function might vary across types of demands [@Arulpragasam2018; @Biaaszek2017; @Chong2017; @Hartmann2013; @Klein-Flugge2016; @Kool2018; @Prevost2010] and the degree to which discount rates correlate across cost domains [@Seaman2018; @Westbrook2013]. A limitation of the discounting approach is that it does not directly probe how aspects of the decision context, such as opportunity costs, might influence cost evaluation, and constrains the understanding of behavioral demands to one in which such demands necessarily reduce the value of prospects.

  Single choice foraging paradigms, in which decision makers choose to accept or reject individually presented prospects, provide a natural way to investigate contextual influences on cost evaluation [@Constantino2015; @Garrett2020; @Krebs1977; @Mobbs2018; @Stephens1986]. Such paradigms require weighing a prospective reward against the perceived opportunity cost of time. Specifically, foragers must decide whether to spend their limited time by harvesting a currently available resource or by seeking out more profitable (or less costly) alternatives instead. This could involve, for example, deciding when to leave a patch of depleting richness (e.g. when to quit a job that provides diminishing opportunities for growth), or choosing whether to pursue or forgo offers presented serially (e.g. whether to accept your first job offer or wait for new potential ones). Here we focus on this latter type of foraging, traditionally called *prey selection* [@Krebs1977], as it provides a means to examine how choices to accept or reject individual prospects are influenced by contextual knowledge about other potentially obtainable alternatives in the same environment.

@@

```{r General prop & earnings plots (Fig. 2), eval = !figsEnd, fig.width = 10, fig.height = 4, fig.align = "center", fig.cap = "Overall choice behavior in Experiment 1. A: Proportion accepted per cost. B: Total number of dollars earned by each group by the end of the experiment (not including the participation reward). C: The relationship between proportion accepted and total earned. Consistent with the foraging design, participants who over and under accepted earned the least. These results suggest that people were willing to pursue cognitively effortful actions at the expense of earnings."}
# plots generated in 16.2.1. and 16.1.3.
# propPlot <- (prop1 | earnings1 | earnFitplot_btw) + 
#     plot_annotation(tag_levels = 'A') &
#     theme(plot.tag = element_text(size = 20, face = "bold"))
# ggsave("./Images/FIG2_props.png", propPlot, device = "png", height = 4, width = 10)

@@

  Stability was defined as the maintenance of similar acceptance rates throughout the experimental session. The scatterplot in Fig. 4 (bottom-left and bottom-right) shows the overall proportion accepted by each participant before and after the midpoint of the session. Paired permutations showed no change in acceptance rates for the Wait and Easy groups (*p* > 0.05), but we found significantly lower acceptance rates in the second half of the session for the Physical group (*p* `r report_p(prepost_btw$Perms[1])`, Cohen's D = `r round(prepost_btw$CohenD[1], digits = 2)`) and the Cognitive group (*p* `r report_p(prepost_btw$Perms[2])`, Cohen's D = `r round(prepost_btw$CohenD[2], digits = 2)`). Fig. 5 (left) shows that the reduction in acceptance rates in effort groups over time was steeper for longer handling times, a hallmark of effortful demands [@Treadway2009]. We then replicated the mixed-effects logistic regression from the previous section separately for each half of the session, which confirmed that differences among demands were stable (Fig. 5, right). Together, the consistency and stability of behavior suggested that decision makers had a clear representation of the environment.

@@

  The original between-subjects task from Experiment 1 was modified so that participants foraged for cost types in addition to rewards. Every participant experienced all forms of cost (delay, physical effort, and cognitive effort). There were two block types, in which delay trials were interspersed with either physical-effort or cognitive-effort trials. This setup prevented participants from having to rapidly switch between response modalities across trials (i.e. keyboard and handgrip). Each block combination (Physical/Wait and Cognitive/Wait) was experienced three times by each participant in 7-minute-long, interleaved blocks, for a total session duration of 42 minutes. Half of the participants experienced a block sequence that started with the Physical/Wait block and the other half experienced a sequence that started with the Cognitive/Wait block. Timing parameters were held fixed throughout the experiment and were matched to the middle condition from Experiment 1 (10 s handling time and 6 s travel time). Participants were informed that this timing combination would be constant prior to beginning the experiment. The upcoming effort type was disclosed at the beginning of each block, and reward/cost offers were displayed at the start of each trial (e.g. “8 cents for physical effort”). Unlike the previous experiment, participants could express their decision to quit by pressing the spacebar during the offer window (not only during the handling time), in which case the travel time would begin immediately at the end of the 2 s offer window. Participants were trained in each type of demand until they reached the same criteria as in Experiment 1.

@@

  In Experiment 2, we tested decision makers' preferences for cognitive effort or physical effort relative to unfilled delay delay. We modified the prey-selection foraging task, so each participant experienced every type of demand in blocks that interleaved wait trials with either physical or cognitive effort (with a fixed 10 s handling time and 6 s travel time, to allow for comparisons between experiments). We separately analyzed Wait trials from the two block types, labeling the two sets of Wait trials according to the type of effort with which they were paired (“Wait-C” for cognitive, and “Wait-P” for physical). We hypothesized that 1) reward offer amounts would influence decisions in the same way as in Experiment 1; 2) interleaving effort and delay would produce higher acceptance rates for wait trials than for cognitive or physical effort trials, in contrast to the pattern observed in Experiment 1. and 3) that choice patterns would remain stable over time and consistent within trials.

@@

```{r iBIC and parameter comparison (Fig. 9), eval = !figsEnd, fig.align = "center", fig.height = 5, fig.width = 7, fig.cap = "Computational modeling results. A: iBIC values for four candidate models of choice behavior for both experiments. Lower values indicate better fit. B: Estimates of the bias parameter c~demand~ for participants in the Physical, Cognitive, and Wait cost conditions of Experiment 1. Negative values suggest the Cognitive group acted as if cognitive demands attenuated the perceived opportunity cost of the delay. C: Estimates of other model parameters indicated consistent foraging behavior across conditions and experiments. D: Synthetic behavior for Experiment 2 generated by the winning model. Error bars correspond to the empirical mean acceptance rate ± SEM per cost type, whereas lines and shaded areas show the model-generated mean acceptance rate and its SEM, respectively. The model's behavior recapitulated the initial apparent preference for cognitive effort and the eventual convergence of acceptance rates by the end of the session."}
# modelPlot <- (bicPlot + costPlot) / allFitsPlot
# ggsave("./Images/FIG9_modelresult_part.png", modelPlot, device = "png", height = 8, width = 11)
include_graphics("./Images/FIG9_modelresults.png")

@@

  The observed choice dynamics ruled out a number of candidate explanations. For instance, if the subjective costs and benefits of effort or delay were intrinsic, we would not have expected to observe the gradual convergence in acceptances when all three forms of demand were intermixed in the same environment. This line of reasoning also speaks against a direct role of boredom in our results. The aversiveness of boredom can lead people to attempt any available activity in order to avoid it [@westgate2020; @Wilson2014]. In our paradigm, such a tendency would have been expected to lead to a persistent preference for either effortful condition. Lastly, the gradual convergence of acceptance rates across cost conditions  rules out the possibility that our sample was mostly composed of demand seekers. Some individuals express a strong need for cognition, which can bias studies of mental labor [@Cacioppo1996; @Sayali2019]. Even though the prevalence of this trait in our college-level population could have driven the preferences for cognitive effort observed in both experiments, we would have expected such individuals to maintain this preference in the face of alternatives. Another unlikely candidate is fatigue. Even though participants decreased their acceptance rates for effort in both experiments, we would have expected participants in Experiment 2 to increase engagement with Wait trials, substituting Wait trials for effort trials in order to preserve their earnings. Our observations are unlikely to be due to the risk of failing to successfully complete effortful trials, as participants remained accurate throughout both experiments. Finally, we cannot conclude that cognitive effort gained value from its rewarding outcomes, as has been proposed in the past [@Eisenberger1996]. This phenomenon would not have been expected to benegit a single type of demand differentially because cognitive effort, physical effort, and unfilled delay were rewarded at the same rate.

@@

```{r New General prop & earnings plots (Fig. 2), eval = figsEnd, fig.width = 10, fig.height = 4, fig.align = "center", fig.cap = "Overall choice behavior in Experiment 1. A: Proportion accepted per cost. B: Total number of dollars earned by each group by the end of the experiment (not including the participation reward). C: The relationship between proportion accepted and total earned. Consistent with the foraging design, participants who over and under accepted earned the least. These results suggest that people were willing to pursue cognitively effortful actions at the expense of earnings."}
# plots generated in 16.2.1. and 16.1.3.
# propPlot <- (prop1 | earnings1 | earnFitplot_btw) + 
#     plot_annotation(tag_levels = 'A') &
#     theme(plot.tag = element_text(size = 20, face = "bold"))
# ggsave("./Images/FIG2_props.png", propPlot, device = "png", height = 4, width = 10)
include_graphics("./Images/FIG2_props.png")

@@

```{r New BIC and parameter comparison (Fig. 9), eval = figsEnd, fig.align = "center", fig.height = 5, fig.width = 7, fig.cap = "Computational modeling results. A: iBIC values for four candidate models of choice behavior for both experiments. Lower values indicate better fit. B: Estimates of the bias parameter c~demand~ for participants in the Physical, Cognitive, and Wait cost conditions of Experiment 1. Negative values suggest the Cognitive group acted as if cognitive demands attenuated the perceived opportunity cost of the delay. C: Estimates of other model parameters indicated consistent foraging behavior across conditions and experiments. D: Synthetic behavior for Experiment 2 generated by the winning model. Error bars correspond to the empirical mean acceptance rate ± SEM per cost type, whereas lines and shaded areas show the model-generated mean acceptance rate and its SEM, respectively. The model's behavior recapitulated the initial apparent preference for cognitive effort and the eventual convergence of acceptance rates by the end of the session."}
# modelPlot <- (bicPlot + costPlot) / allFitsPlot
# ggsave("./Images/FIG9_modelresult_part.png", modelPlot, device = "png", height = 8, width = 11)
include_graphics("./Images/FIG9_modelresults.png")
```
